---
title: "Econometrics II - Assignment 2"
author: Floris Holstege, Stanislav Avdeev
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Packages required for subsequent analysis. P_load ensures these will be installed and loaded. 
if (!require("pacman")) install.packages("pacman")
pacman::p_load(rio,
               stargazer,
               plm,
               parameters)

# load dataset
df <- import("Data/NLSY2000RC_V2.csv")
df$EARNINGS <- log(df$EARNINGS)


# covariance matrix with White SE
vcovWhite <- function(x) vcovHC(x, method="white1")

```

## 1

```{r include=FALSE, eval=FALSE, echo=FALSE, results='hide'}
# Run poooled OLS including and excluding ASVABC
lm_pooled_no_ab <- plm(EARNINGS ~ S + AGE + AGESQ + ETHBLACK + URBAN + REGNE + REGNC + REGW,
                    data = df, model = "pooling", index = "ID")
lm_pooled_ab <- plm(EARNINGS ~ S + ASVABC + AGE + AGESQ + ETHBLACK + URBAN + REGNE + REGNC + REGW,
                    data = df, model = "pooling", index = "ID")

stargazer(lm_pooled_no_ab, lm_pooled_ab,
          keep.stat=c("n","adj.rsq"),
          title="OLS pooled model with and without ability variable")

# we should use robust s.e.
# and change variable names to decent ones
# add citation
```

When we do not account for ability in the model specification, returns to one year of education are higher, as expected. Higher ability students tend to get more education, thus, they tend to get higher earnings.

From now on we are going to include the ability variable in all model specifications. For the sake of saving the space we don't report all coefficients in tables unless they are important for a specific question.

\begin{table}[!htbp] \centering 
  \caption{OLS pooled model with and without ability variable} 
  \label{} 
\begin{tabular}{@{\extracolsep{5pt}}lcc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{2}{c}{\textit{Dependent variable:}} \\ 
\cline{2-3} 
\\[-1.8ex] & \multicolumn{2}{c}{EARNINGS} \\ 
\\[-1.8ex] & (1) & (2)\\ 
\hline \\[-1.8ex] 
 S & 0.070$^{***}$ & 0.048$^{***}$ \\ 
  & (0.001) & (0.001) \\ 
  & & \\ 
 ASVABC &  & 0.011$^{***}$ \\ 
  &  & (0.0003) \\ 
  & & \\ 
 Constant & $-$0.079 & $-$0.386$^{***}$ \\ 
  & (0.051) & (0.051) \\ 
  & & \\ 
\hline \\[-1.8ex] 
Observations & 40,043 & 40,043 \\ 
Adjusted R$^{2}$ & 0.292 & 0.313 \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Note:}  & \multicolumn{2}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
\end{tabular} 
\end{table} 

## 2


```{r include=FALSE, eval=FALSE, echo=FALSE, results='hide'}
# Include interaction between schooling and ethnicity & run two models for two races separately
lm_inter <- plm(EARNINGS ~ S + S*ETHBLACK + ASVABC + AGE + AGESQ + ETHBLACK + URBAN + REGNE + REGNC + REGW,
                    data = df, model = "pooling", index = "ID")
lm_white <- plm(EARNINGS ~ S + ASVABC + AGE + AGESQ + URBAN + REGNE + REGNC + REGW,
                    data = df[df$ETHBLACK == 0,], model = "pooling", index = "ID")
lm_black <- plm(EARNINGS ~ S + ASVABC + AGE + AGESQ + URBAN + REGNE + REGNC + REGW,
                    data = df[df$ETHBLACK == 1,], model = "pooling", index = "ID")
stargazer(lm_inter, lm_white, lm_black,
          keep.stat=c("n","adj.rsq"),
          title="OLS pooled model with heterogeneous effects by ethnicity")
```

As we can see, there is a statistically significant difference between returns to education by ethnicity. Black workers tend to get higher returns.

\begin{table}[!htbp] \centering 
  \caption{OLS pooled model with heterogeneous effects by ethnicity} 
  \label{} 
\begin{tabular}{@{\extracolsep{5pt}}lccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{3}{c}{\textit{Dependent variable:}} \\ 
\cline{2-4} 
\\[-1.8ex] & \multicolumn{3}{c}{EARNINGS} \\ 
\\[-1.8ex] & (1) & (2) & (3)\\ 
\hline \\[-1.8ex] 
 S & 0.046$^{***}$ & 0.046$^{***}$ & 0.061$^{***}$ \\ 
  & (0.001) & (0.001) & (0.003) \\ 
  & & & \\ 
 S:ETHBLACK & 0.016$^{***}$ &  &  \\ 
  & (0.003) &  &  \\ 
  & & & \\ 
 Constant & $-$0.370$^{***}$ & $-$0.437$^{***}$ & 0.038 \\ 
  & (0.051) & (0.054) & (0.140) \\ 
  & & & \\ 
\hline \\[-1.8ex] 
Observations & 40,043 & 35,223 & 4,820 \\ 
Adjusted R$^{2}$ & 0.313 & 0.299 & 0.314 \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Note:}  & \multicolumn{3}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
\end{tabular} 
\end{table} 

## 3 

Once we use a panel model with random effects, there seems to be no difference returns to education by ethnicity.

```{r include=FALSE, eval=FALSE, echo=FALSE, results='hide'}
# Run random effect models with heteregoneity
lm_inter_rand <- plm(EARNINGS ~ S + S*ETHBLACK + ASVABC + AGE + AGESQ + ETHBLACK + URBAN + REGNE + REGNC + REGW,
                    data = df, model = "random", index = "ID")
lm_white_rand <- plm(EARNINGS ~ S + ASVABC + AGE + AGESQ + URBAN + REGNE + REGNC + REGW,
                    data = df[df$ETHBLACK == 0,], model = "random", index = "ID")
lm_black_rand <- plm(EARNINGS ~ S + ASVABC + AGE + AGESQ + URBAN + REGNE + REGNC + REGW,
                    data = df[df$ETHBLACK == 1,], model = "random", index = "ID")

stargazer(lm_inter_rand, lm_white_rand, lm_black_rand,
          keep.stat=c("n","adj.rsq"),
          title="Panel regressions with random effects")
```

\begin{table}[!htbp] \centering 
  \caption{Panel regressions with random effects} 
  \label{} 
\begin{tabular}{@{\extracolsep{5pt}}lccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{3}{c}{\textit{Dependent variable:}} \\ 
\cline{2-4} 
\\[-1.8ex] & \multicolumn{3}{c}{EARNINGS} \\ 
\\[-1.8ex] & (1) & (2) & (3)\\ 
\hline \\[-1.8ex] 
 S & 0.051$^{***}$ & 0.049$^{***}$ & 0.052$^{***}$ \\ 
  & (0.002) & (0.002) & (0.006) \\ 
  & & & \\ 
 S:ETHBLACK & $-$0.004 &  &  \\ 
  & (0.006) &  &  \\ 
  & & & \\ 
 Constant & $-$0.452$^{***}$ & $-$0.467$^{***}$ & $-$0.223$^{*}$ \\ 
  & (0.048) & (0.050) & (0.122) \\ 
  & & & \\ 
\hline \\[-1.8ex] 
Observations & 40,043 & 35,223 & 4,820 \\ 
Adjusted R$^{2}$ & 0.366 & 0.371 & 0.322 \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Note:}  & \multicolumn{3}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
\end{tabular} 
\end{table} 


## 4

## 5

```{r include=FALSE, eval=FALSE, echo=FALSE, results='hide'}
# Run fixed effect models with heteregoneity
lm_inter_fix <- plm(EARNINGS ~ S + S*ETHBLACK + ASVABC + AGE + AGESQ + ETHBLACK + URBAN + REGNE + REGNC + REGW,
                    data = df, model = "within", index = "ID")
lm_white_fix <- plm(EARNINGS ~ S + ASVABC + AGE + AGESQ + URBAN + REGNE + REGNC + REGW,
                    data = df[df$ETHBLACK == 0,], model = "within", index = "ID")
lm_black_fix <- plm(EARNINGS ~ S + ASVABC + AGE + AGESQ + URBAN + REGNE + REGNC + REGW,
                    data = df[df$ETHBLACK == 1,], model = "within", index = "ID")

stargazer(lm_inter_fix, lm_white_fix, lm_black_fix,
          keep.stat=c("n","adj.rsq"),
          title="Panel regressions with fixed effects")
```

\begin{table}[!htbp] \centering 
  \caption{Panel regressions with fixed effects} 
  \label{} 
\begin{tabular}{@{\extracolsep{5pt}}lccc} 
\\[-1.8ex]\hline 
\hline \\[-1.8ex] 
 & \multicolumn{3}{c}{\textit{Dependent variable:}} \\ 
\cline{2-4} 
\\[-1.8ex] & \multicolumn{3}{c}{EARNINGS} \\ 
\\[-1.8ex] & (1) & (2) & (3)\\ 
\hline \\[-1.8ex] 
 S & 0.053$^{***}$ & 0.049$^{***}$ & 0.027$^{**}$ \\ 
  & (0.004) & (0.004) & (0.011) \\ 
  & & & \\ 
 S:ETHBLACK & $-$0.062$^{***}$ &  &  \\ 
  & (0.012) &  &  \\ 
  & & & \\ 
\hline \\[-1.8ex] 
Observations & 40,043 & 35,223 & 4,820 \\ 
Adjusted R$^{2}$ & 0.152 & 0.165 & 0.047 \\ 
\hline 
\hline \\[-1.8ex] 
\textit{Note:}  & \multicolumn{3}{r}{$^{*}$p$<$0.1; $^{**}$p$<$0.05; $^{***}$p$<$0.01} \\ 
\end{tabular} 
\end{table} 

## 6 

To decide between fixed or random effects we run a Hausman test. The p-value is highly significant, thus, it's preferred to use fixed effects.

```{r include=FALSE, eval=FALSE, echo=FALSE, results='hide'}
phtest(lm_inter_fix, lm_inter_rand)
# should we run any other tests?
```

## 7

```{r include=FALSE, eval=FALSE, echo=FALSE, results='hide'}
# Create within-mean variables
df <- cbind(df, demean(df, select = c("AGE", "S", "URBAN", "REGNE", "REGNC", "REGW", "AGESQ"), 
                      group = "ID"))
df <- df[,c(1:20)]

# Run Mundlak model
gls_mundlak <- pggls(EARNINGS ~ S + S*ETHBLACK + ASVABC + AGE + AGESQ + ETHBLACK + URBAN + REGNE + REGNC + REGW + S_between + AGE_between + AGESQ_between + URBAN_between + REGNE_between + REGNC_between + REGW_between,
                    data = df, model = "pooling", index = "ID")
summary(gls_mundlak)
```

## 8

## 9

```{r include=FALSE, eval=FALSE, echo=FALSE, results='hide'}
# Apply Verbeek and Nijman test
freq <- table(df$ID) 
freq <- as.data.frame(freq)
  
freq$Freq <- ifelse(freq$Freq > 4, freq$Freq, NA)
freq <- na.omit(freq)
```

